A finite-state morphological transducer for Kyrgyz

### 2012-03-14 paper/paper.tex
misc

\usepackage{fontspec}
\defaultfontfeatures{PunctuationSpace=3,Scale=MatchLowercase,Mapping=tex-text}
\newfontfeature{IPA}{+mgrk}
\setromanfont[IPA]{FreeSerif}
%\setromanfont[IPA]{Times New Roman}
%\renewcommand{\sfdefault}{phv}
%\renewcommand{\rmdefault}{ptm}
%\renewcommand{\ttdefault}{pcr}
\usepackage[small,bf]{caption}

@@

\section{Background}
\subsection{The Kyrgyz language}
Kyrgyz (written ‹кыргыз тили› or ‹\textuighur{قىرعىز تىلى}›, pronounced [qɯɾʁɯz tilí]) alternatively written ``Kirghiz'' or ``Kirgiz'') is a Turkic language spoken in Kyrgyzstan, China, Tajikistan, and Uzbekistan.  Its classification within Turkic remains problematic—it appears to alternatively belong to the Kypchak (Northwestern) branch and to the South Siberian (Northeastern) branch.  The Turkic varieties phonetically and phonologically most similar to Kyrgyz are the southern dialects of Altay, though Kyrgyz shows strong parallels to Kazakh that these varieties lack, especially in its Talas dialects.  In southern varieties of Kyrgyz there are also many similarities to Uzbek that other dialects lack.

@@

The analyser is designed to be used with the longest-match left-to-right (LRLM) tokenisation strategy, as described in \cite{garrido-alenda02}. Thus when more than one `word' needs to be output, they are joined with the \texttt{+} symbol, which is reserved for joining two `words' in one analysis.  For example, `өнүктүрөбүзбү ?' \eng{will we develop it?} is analysed as 
өнүк{\tt {\small <v><tv><caus><aor><p1><pl>+}}бы{\tt {\small <qst>}}, which is two words: `өнүк' (verb, transitive, causative, aorist, first person, plural) and `бы' (question clitic).  Likewise, `келатсаң' \eng{if you come} кел{\tt {\small <v><iv><prt\_impf>}}+жат\texttt{{\small <vaux><gna\_cnd><p2><sg>}} is analysed as two words: `кел' (verb, intransitive, imperfect participle), and жат (auxiliary, conditional verbal adverb, second person, singular).

@@

Here, the underlying /D/ is always realised as [д] when after a voiced segment (including a vowel), but is realised as [т] after an unvoiced consonant, while underlying /G/ is always realised as [г]\footnote{‹г› is pronounced [ɣ] in the context of front vowels and [ʁ] in the context of back vowels; it also has realisations of [ɡ] and [ɢ] after nasals.} after a voiced segment, but is realised as [к]\footnote{‹к› is realised as [q] in the context of back vowels and [k] in the presence of front vowels.} after voiceless consonants.

@@

\subsection{й+vowel letters}
\label{sec:yot}
In Kyrgyz, there are a series of letters which represent /й/ plus a vowel: /я, е, ё, ю/ represent /йа, йэ, йо, йу/ respectively.  However, /э/ is also represented as /е/ when short and after a consonant (i.e., it is only represented as /э/ when long—/ээ/—or when word-initially).  These ``yoticised'' vowels proved difficult to work.  For example, /бой+(S)I/ \eng{length}, given normal vowel harmony rules, would surface as [бойу].  However, the correct form is [бою].  Distinct and separate vowel harmony rules had to be created for all vowel archiphonemes in post-vocalic contexts, e.g. here to turn \{I\} into [ю] after /й/.  An additional rule had to delete the underlying /й/ before yoticised vowel letters (so that e.g. [бойю] was not output).  Because there are too levels to the phonology in twolc and it is otherwise not linear, attention had to be paid to the fact that e.g., the distinct vowel harmony rules for post-/й/ contexts recognised that the /й/ would not appear on the second [surface] level.

>>>>

\usepackage{fontspec}
\defaultfontfeatures{PunctuationSpace=3,Scale=MatchLowercase,Mapping=tex-text}
\newfontfeature{IPA}{+mgrk}
%\setromanfont[IPA]{FreeSerif}
%\setromanfont[IPA]{Times New Roman}
%\setromanfont[IPA]{NimbusRomNo9L-Regu}
\setromanfont{Times New Roman}
%\renewcommand{\sfdefault}{phv}
%\renewcommand{\rmdefault}{ptm}
%\renewcommand{\ttdefault}{pcr}
\usepackage[small,bf]{caption}
%\newfontfamily\qipa[IPA]{NimbusRomNo9L-Regu}
\newfontfamily\qipa[IPA,Scale=MatchLowercase]{FreeSerif}

@@

\section{Background}
\subsection{The Kyrgyz language}
Kyrgyz (written ‹кыргыз тили› or ‹\textuighur{قىرعىز تىلى}›, pronounced [{\qipa qɯɾʁɯz tilí}]) alternatively written ``Kirghiz'' or ``Kirgiz'') is a Turkic language spoken in Kyrgyzstan, China, Tajikistan, and Uzbekistan.  Its classification within Turkic remains problematic—it appears to alternatively belong to the Kypchak (Northwestern) branch and to the South Siberian (Northeastern) branch.  The Turkic varieties phonetically and phonologically most similar to Kyrgyz are the southern dialects of Altay, though Kyrgyz shows strong parallels to Kazakh that these varieties lack, especially in its Talas dialects.  In southern varieties of Kyrgyz there are also many similarities to Uzbek that other dialects lack.

@@

The analyser is designed to be used with the longest-match left-to-right (LRLM) tokenisation strategy, as described in \cite{garrido-alenda02}. Thus when more than one `word' needs to be output, they are joined with the \texttt{+} symbol, which is reserved for joining two `words' in one analysis.  For example, `өнүктүрөбүзбү ?' \eng{will we develop it?} is analysed as 
өнүк{\tt {\small <v><tv><caus><aor><p1><pl>+}}бы{\tt {\small <qst>}}, which is two words: `өнүк' (verb, transitive, causative, aorist, first person, plural) and `бы' (question clitic).  Likewise, `келатсаң' \eng{if you come} кел{\tt {\small <v><iv><prt\_impf>}}+жат\texttt{{\small <vaux><gna\_cnd><p2><sg>}} is analysed as two words: `кел' (verb, intransitive, imperfect participle), and `жат' (auxiliary, conditional verbal adverb, second person, singular).

@@

Here, the underlying /D/ is always realised as [д] when after a voiced segment (including a vowel), but is realised as [т] after an unvoiced consonant, while underlying /G/ is always realised as [г]\footnote{‹г› is pronounced [{\qipa ɣ}] in the context of front vowels and [{\qipa ʁ}] in the context of back vowels; it also has realisations of [{\qipa ɡ}] and [{\qipa ɢ}] after nasals.} after a voiced segment, but is realised as [к]\footnote{‹к› is realised as [{\qipa q}] in the context of back vowels and [{\qipa k}] in the presence of front vowels.} after voiceless consonants.

@@

\subsection{й+vowel letters}
\label{sec:yot}
In Kyrgyz, there are a series of letters which represent /й/ plus a vowel: /я, е, ё, ю/ represent /йа, йэ, йо, йу/ respectively.  However, /э/ is also represented as /е/ when short and after a consonant (i.e., it is only represented as /э/ when long—/ээ/—or when word-initially).  These ``yoticised'' vowels proved difficult to work with.  For example, /бой+(S)I/ \eng{length}, given normal vowel harmony rules, would surface as [бойу].  However, the correct orthographic form is [бою].  Distinct and separate vowel harmony rules had to be created for all vowel archiphonemes in post-vocalic contexts, e.g.\ here to turn \{I\} into [ю] after /й/.  An additional rule had to delete the underlying /й/ before yoticised vowel letters (so that e.g.\ [бойю] was not output).  Because there are two levels to the phonology in twolc and it is otherwise not linear, attention had to be paid to the fact that e.g., the distinct vowel harmony rules for post-/й/ contexts were aware that the /й/ would not appear on the second [surface] level.
